---
title: "Class11"
author: "Joel Kosareff"
format: gfm
---
>Q13: Read this file into R and determine the sample size for each genotype and their
corresponding median expression levels for each of these genotypes.

```{r}
Data <- read.table("https://bioboot.github.io/bggn213_W19/class-material/rs8067378_ENSG00000172057.6.txt")
View(Data)
```

```{r}
summary(Data)
```
```{r}
library(stringr)
sum(str_count(Data$geno, "A/A"))
sum(str_count(Data$geno, "A/G"))
sum(str_count(Data$geno, "G/G"))
```

There are 108 A/A, 233 A/G and 121 G/G

```{r}
median(Data$exp[str_detect(Data$geno, "A/A")])
median(Data$exp[str_detect(Data$geno, "A/G")])
median(Data$exp[str_detect(Data$geno, "G/G")])

```
Median of A/A: 31.25
Median of A/G: 25.06
Median of G/G: 20.07

>Q14: Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3?

```{r}
library(ggplot2)
```

```{r}
ggplot(Data) + aes(geno, exp) + geom_boxplot()

```

It would appear that Asthma is expressed higher in the A/A population than it is in the G/G population. This proposes that the SNP does in fact have an effect on the expression of ORMDL3 