---
title: "Class05"
author: "Joel Kosareff"
format: gfm
---

Install the package ggplot2
```{r}
#install.packages("ggplot2")
```

Any time I want to use this package I need to load it.
```{r}
library(ggplot2)
```

```{r}
View(cars)
```

A quick base R plot - not in ggplot2
```{r}
plot(cars)
```

Our First ggplot. We need data + aes + geoms
```{r}
ggplot(data = cars) + aes(x = speed, y = dist) + geom_point()
```

```{r}
p <- ggplot(data = cars) + aes(x = speed, y = dist) + geom_point()
```

Add a line geometry with line_geom()
```{r}
p + geom_line()
```


Add a trend line close to the data
```{r}
p + geom_smooth()
```

```{r}
p + geom_smooth(method="lm", se = FALSE)
```

```{r}
p + labs(title="Speed and Stopping Distances of Cars", x = "Speed (MPH)", y = "Stopping Distance (ft)", subtitle = "Your informative subtitle text here", caption = "Dataset: 'cars'") + geom_smooth(method="lm", se = FALSE)
```


Read in our drug expression data
```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```

Q. How many genes are in this dataset?
```{r}
nrow(genes)
```
```{r}
ncol(genes)
```

How many Up regulated genes are there?
```{r}
table(genes$State)
```

What fraction of genes are up regulated?
```{r}
round(table(genes$State)/nrow(genes) * 100, 2)
```

Lets make a first plot attempt 
```{r}
g <- ggplot(data = genes) + aes(x = Condition1, y = Condition2, col = State) +   geom_point()
g
```


Add some color
```{r}
g + scale_color_manual(values=c("blue","gray","red")) + labs(title = "Gene Expression", x = "Control (no drug)") + theme_bw()
```

