---
title: "Class13"
author: "Joel Kosareff"
format: gfm
---

## Differential Expression Analysis 

```{r}
library(DESeq2)
```

Now load the data
```{r}
metaFile <- "GSE37704_metadata.csv"
countFile <- "GSE37704_featurecounts.csv"

```

```{r}
colData = read.csv(metaFile, row.names = 1)
head(colData)
```

```{r}
countData = read.csv(countFile, row.names=1)
head(countData)
```

We need to remove the first column in count data

```{r}
countData <- as.matrix(countData[,-1])
head(countData)
```

Next lets remove the zero entries 

```{r}
countData = countData[rowSums(countData[])>0, ]
head(countData)
```

Now we set up the DESeq object

```{r}
dds = DESeqDataSetFromMatrix(countData=countData,
                             colData=colData,
                             design=~condition)
dds = DESeq(dds)
```

```{r}
dds
```

Next lets get results for the HoxA1 knockdown versus control siRNA

```{r}
res = results(dds, contrast=c("condition", "hoxa1_kd", "control_sirna"))
```

```{r}
summary(res)
```

Time for a volcano plot

```{r}
plot( res$log2FoldChange, -log(res$padj) )
```

```{r}
# Make a color vector for all genes
mycols <- rep("gray", nrow(res) )

# Color red the genes with absolute fold change above 2
mycols[ abs(res$log2FoldChange) > 2 ] <- "red"

# Color blue those with adjusted p-value less than 0.01
#  and absolute fold change more than 2
inds <- (res$pvalue < 0.01) & (abs(res$log2FoldChange) > 2 )
mycols[ inds ] <- "blue"

plot( res$log2FoldChange, -log(res$padj), col=mycols, xlab="Log2(FoldChange)", ylab="-Log(P-value)" )
```

```{r}
library("AnnotationDbi")
library("org.Hs.eg.db")

columns(org.Hs.eg.db)

res$symbol = mapIds(org.Hs.eg.db,
                    keys=row.names(res), 
                    keytype="ENSEMBL",
                    column="SYMBOL",
                    multiVals="first")

res$entrez = mapIds(org.Hs.eg.db,
                    keys=row.names(res),
                    keytype="ENSEMBL",
                    column="ENTREZID",
                    multiVals="first")

res$name =   mapIds(org.Hs.eg.db,
                    keys=row.names(res),
                    keytype="ENSEMBL",
                    column="GENENAME",
                    multiVals="first")

head(res, 10)
```

Now lets order by p-value

```{r}
res = res[order(res$pvalue),]
write.csv(res[],file ="deseq_results.csv")
```


## Pathway Analysis

We are going to use the Gage and Pathview packages for pathway analysis 

```{r}
library(pathview)
```

```{r}
library(gage)
library(gageData)

data(kegg.sets.hs)
data(sigmet.idx.hs)

kegg.sets.hs = kegg.sets.hs[sigmet.idx.hs]

head(kegg.sets.hs, 3)
```
The main gage() function requires a named vector of fold changes, where the names of the values are the Entrez gene IDs.

Note that we used the mapIDs() function above to obtain Entrez gene IDs (stored in res$entrez) and we have the fold change results from DESeq2 analysis (stored in res$log2FoldChange).

```{r}
foldchanges = res$log2FoldChange
names(foldchanges) = res$entrez
head(foldchanges)
```

Now lets run gage analysis

```{r}
keggres = gage(foldchanges, gsets=kegg.sets.hs)
```

```{r}
attributes(keggres)
```

Lets look at the first few down (less) pathway results:

```{r}
head(keggres$less)
```

Now, let's try out the pathview() function from the pathview package to make a pathway plot with our RNA-Seq expression results shown in color.
To begin with lets manually supply a pathway.id (namely the first part of the "hsa04110 Cell cycle") that we could see from the print out above.

```{r}
pathview(gene.data=foldchanges, pathway.id="hsa04110")
```

We can also generate a pdf

```{r}
pathview(gene.data=foldchanges, pathway.id="hsa04110", kegg.native=FALSE)
```

Now, let's process our results a bit more to automagicaly pull out the top 5 upregulated pathways, then further process that just to get the pathway IDs needed by the pathview() function. We'll use these KEGG pathway IDs for pathview plotting below.


```{r}
keggrespathways <- rownames(keggres$greater)[1:5]
keggresids = substr(keggrespathways, start=1, stop=8)
keggresids
```

Finally, lets pass these IDs in keggresids to the pathview() function to draw plots for all the top 5 pathways.

```{r}
pathview(gene.data=foldchanges, pathway.id=keggresids, species="hsa")
```

And the down regualted pathways

```{r}
keggrespathways <- rownames(keggres$less)[1:5]
keggresids = substr(keggrespathways, start=1, stop=8)
keggresids
```

```{r}
pathview(gene.data=foldchanges, pathway.id=keggresids, species="hsa")
```


## Gene Ontology

We can also do a similar procedure with gene ontology. Similar to above, go.sets.hs has all GO terms. go.subs.hs is a named list containing indexes for the BP, CC, and MF ontologies. Let’s focus on BP (a.k.a Biological Process) here.


```{r}
data(go.sets.hs)
data(go.subs.hs)

gobpsets = go.sets.hs[go.subs.hs$BP]

gobpres = gage(foldchanges, gsets=gobpsets, same.dir=TRUE)

lapply(gobpres, head)
```



## Reactome Analysis

Let's now conduct over-representation enrichment analysis and pathway-topology analysis with Reactome using the previous list of significant genes generated from our differential expression results above.

First, Using R, output the list of significant genes at the 0.05 level as a plain text file:

```{r}
sig_genes <- res[res$padj <= 0.05 & !is.na(res$padj), "symbol"]
print(paste("Total number of significant genes:", length(sig_genes)))
```

```{r}
write.table(sig_genes, file="significant_genes.txt", row.names=FALSE, col.names=FALSE, quote=FALSE)
```


The endosomal/vacuolar pathway has the most significant p-value 
