---
title: "Class10"
author: "Joel Kosareff"
format: gfm
---

Lets read in the candy file
```{r}
candy_file <- "candy-data.csv"

candy = read.csv(candy_file, row.names = 1)
```

>Q1. How many different candy types are in this dataset?

```{r}
nrow(candy)
```
85    

>Q2. How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)
```
38

Lets see how popular twix is 
```{r}
candy["Twix", ]$winpercent
```



>Q3. What is your favorite candy in the dataset and what is it’s winpercent value?

```{r}
candy["100 Grand", ]$winpercent
```


>Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy["Kit Kat", ]$winpercent
```


>Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars", ]$winpercent
```

```{r}
#install.packages("skimr")
```

```{r}
library("skimr")
skim(candy)
```


>Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

Winpercent is on a different scale 

>Q7. What do you think a zero and one represent for the candy$chocolate column?

Whether or not the candy is a chocolate based candy

>Q8. Plot a histogram of winpercent values

Histogram

```{r}
hist(candy$winpercent)
```


>Q9. Is the distribution of winpercent values symmetrical?

The distribution of winpercent is not symmetrical

>Q10. Is the center of the distribution above or below 50%?

The center of the distribution is below 50
>Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
mean(candy$winpercent[as.logical(candy$chocolate)])
mean(candy$winpercent[as.logical(candy$fruity)])
```
On average chocolate candy is ranked higher

>Q12. Is this difference statistically significant?

```{r}
t.test(candy$chocolate, candy$fruity)
```

This difference is not statistically significant 

```{r}
cd <- (candy[order(candy$winpercent),])
head(cd)
```

>Q13. What are the five least liked candy types in this set?

Nik L Nip, Boston Baked Beans, Chiclets, Super Bubble, Jawbusters

>Q14. What are the top 5 all time favorite candy types out of this set?

Reeses peanut butter cump, reeses miniatures, twix, kit kat, snickers

>Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}
library(ggplot2)

ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) + geom_col()
```

```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"
```

```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) 
```

>Q17. What is the worst ranked chocolate candy?

Charleston Chew

>Q18. What is the best ranked fruity candy?

Starburst

What about value for money? What is the the best candy for the least money? One way to get at this would be to make a plot of winpercent vs the pricepercent variable. The pricepercent variable records the percentile rank of the candy’s price against all the other candies in the dataset. Lower vales are less expensive and high values more expensive.

To this plot we will add text labels so we can more easily identify a given candy. There is a regular geom_label() that comes with ggplot2. However, as there are quite a few candys in our dataset lots of these labels will be overlapping and hard to read. To help with this we can use the geom_text_repel() function from the ggrepel package.

```{r}
library(ggrepel)

# How about a plot of price vs win
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 5)
```


>Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

Reese's Miniatures
>Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```

Nik L Nip, Nestle Smarties, Ring pop, Hershey's Krackel , Heryshey's milk chocolate. Nik L Nip is the least liked

Now that we’ve explored the dataset a little, we’ll see how the variables interact with one another. We’ll use correlation and view the results with the corrplot package to plot a correlation matrix.

```{r}
library(corrplot)
```

```{r}
cij <- cor(candy)
corrplot(cij)
```


>Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

Fruit and Chocolate 
>Q23. Similarly, what two variables are most positively correlated?

bar and chocolate

Let’s apply PCA using the prcom() function to our candy dataset remembering to set the scale=TRUE argument.

```{r}
pca <- prcomp(candy, scale = TRUE)
summary(pca)
```
Now we can plot our main PCA score plot of PC1 vs PC2.
```{r}
plot(pca$x[,1:2])
```

We can change the plotting character and add some color:

```{r}
plot(pca$x[,1:2], col=my_cols, pch=16)
```

We can make a much nicer plot with the ggplot2 package but it is important to note that ggplot works best when you supply an input data.frame that includes a separate column for each of the aesthetics you would like displayed in your final plot. To accomplish this we make a new data.frame here that contains our PCA results with all the rest of our candy data. We will then use this for making plots below

```{r}
# Make a new data-frame with our PCA results and candy data
my_data <- cbind(candy, pca$x[,1:3])
```

```{r}
p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)

p
```

```{r}
library(ggrepel)

p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```

```{r}
library(plotly)
```
```{r}
#I commented out the plotly graph as it was forcing export as html
#ggplotly(p)
```

Let’s finish by taking a quick look at PCA our loadings. Do these make sense to you? Notice the opposite effects of chocolate and fruity and the similar effects of chocolate and bar (i.e. we already know they are correlated).

```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```

>Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

Fruity, hard, and pluribus are noted by PC1. Fruity and hard candies tend to not be related to chocolate and many candies are bought in singles